---
title: "Class_13-Structure_II"
author: "Michael Overton"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will download and clean up the HIV-Pr structure (1HSG) from PDB. We will make separate lists for "protein-only" and "ligand-only" sets.

```{r, echo=FALSE}
library(bio3d)

p.file <- get.pdb("1hsg")
hiv <- read.pdb(p.file)
hiv

```
> Q1)  What is the name of the two non protein resid values in this structure? What does resid correspond to and how would you get a listing of all reside values in this structure?
  > HOH and MK1 
  > resid is short for residue, which corresponds to the identity to each amino acid in the molecule
  > 

```{r, echo=FALSE}
prot <- atom.select(hiv, "protein", value=T)
lig <- atom.select(hiv, "ligand", value=T)
write.pdb(prot, file="1hsg_protein.pdb")
write.pdb(lig, file="1hsg_ligand.pdb")
mk1 <- read.pdb("1hsg_ligand.pdb")
mk1
```

Having processed the protein and ligand structures and run them through Autodocker Vina, we will save the output as a .pbd file
```{r, echo=FALSE}
res <- read.pdb("all.pdbqt", multi=TRUE) 
write.pdb(res, "results.pdb")

```


```{r, echo=FALSE}
res <- read.pdb("all.pdbqt", multi=TRUE)
ori <- read.pdb("1hsg_ligand.pdbqt")
MK1_dist <- rmsd(ori, res)

heavy_res <- atom.select(res, string="noh", value=T)
heavy_ori <- atom.select(ori, string="noh", value=T)

res_matrix <- as.matrix(heavy_res[["xyz"]])
ori_matrix <- as.matrix(heavy_ori[["xyz"]])

MK1_heavy_dist <- c()
for (i in 1:nrow(res_matrix)) {
  d <- rmsd(res_matrix[,i], ori_matrix)
  MK1_heavy_dist <- c(MK1_heavy_dist, d)
}
MK1_heavy_dist <- unlist(MK1_dist)
MK1_heavy_dist
```


```{r, echo=FALSE}

pdb <- read.pdb("1HEL")
modes <- nma(pdb)
plot(modes, sse=pdb)

m7 <- mktrj(modes, mode=7, file="nma_7.pdb")
```


```{r, echo=FALSE}
library("bio3d.view")
view(m7, col=vec2color(rmsf(m7)))

```

