---
title: "Lecture_9-Unsup_Learn"
author: Michael Overton
date: 02.05.2020
output: github_document
---

## K-means clustering

We will try to cluster data using the kmeans() function

```{r}
tmp_data <- c(rnorm(30,-3), rnorm(30,0,0.4), rnorm(30,3))
x <- cbind(x=tmp_data, y=rev(tmp_data))
plot(x)

km <- kmeans(x, 2, 20)
km_3 <- kmeans(x, 3, 50)

```

Use the kmeans() function setting k to 2 and nstart=20
Inspect/print the results
Q. What is in the output object?
```{r}
attributes(km)
```

Q. How many points are in each cluster? 30 and 30
Q. What ‘component’ of your result object details
      - cluster size?
```{r}
km$size
```
      - cluster assignment/membership?
```{r}
km$cluster
```
      - cluster center?    
```{r}
km$centers
```
      
Plot x colored by the kmeans cluster assignment and
      add cluster centers as blue points
```{r}
km_df <- data.frame(x=x[ ,1], y=x[ ,2], k=km$cluster)

plot(x=km_df$x, y=km_df$y, col=km_df$k)
```

```{r echo=F}
km_3_df <- data.frame(x=x[ ,1], y=x[ ,2], k=km_3$cluster)

plot(x=km_3_df$x, y=km_3_df$y, col=km_3_df$k)
points(km_3$centers, pch=15, col="blue", cex=1.5)
```

      
## Hierarchical clustering with distance matrix, hclust, and dendrogram

```{r, echo=F}
dist_x <- dist(x)
h_x <- hclust(d=dist_x)
plot(h_x)
table( cutree(h_x, k=3) )
km_3$size
```

```{r, echo=F}
# Step 1. Generate some example data for clustering 
x2 <- rbind(matrix(rnorm(100, mean=0, sd=0.3), ncol = 2), # c1 
           matrix(rnorm(100, mean=1, sd=0.3), ncol = 2), # c2
           matrix(c(rnorm(50, mean=1, sd=0.3), rnorm(50, mean=0, sd=0.3)), ncol = 2)) # c3
colnames(x2) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x2)
# Step 3. Generate colors for known clusters (just so we can compare to hclust results) 
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x2, col=col)
```

Q. Use the dist(), hclust(), plot() and cutree() functions to return 2 and 3 clusters

```{r, echo=F}
h_x2 <- hclust( d=dist(x2) )
plot(h_x2)
cutree(h_x2, k=2)
cutree(h_x2, k=3)

```


Q. How does this compare to your known 'col' groups?

```{r, echo=F}
plot(x2, col=cutree(h_x2, k=2))
plot(x2, col=cutree(h_x2, k=3))
plot(x2, col=col)

table(cutree(h_x2, k=3))
x2_trueK <- cbind(x2, col)
table(x2_trueK[ ,3])
```


## PCA analysis

```{r, echo=F}
x <- read.csv("UK_foods.csv")
```

Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r, echo=F}
dim(x)
structure(x)
summary(x)
View(x)
head(x, 6)

rownames(x) <- x[,1]
x <- x[,-1]
head(x, 6)
dim(x)
# or

x <- read.csv("UK_foods.csv", row.names = 1)
dim(x)

```

Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
  Ascribing row names inside the read.csv function is shorter and more robust, since the x[ ,-1] argument could be used multiple times and delete useful data from x.

```{r, echo=F}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

Q3: Changing what optional argument in the above barplot() function results in the following plot?
  change "beside=" argument to FALSE

```{r, echo=F}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```


Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

This function produces scatterplots of the values of each food for each pair of countries. If a value lies on the diagonal, it means that the value is equal for both countries.

```{r, echo=F}
pairs(x, col=rainbow(10), pch=16)
```


Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
  Pairwise comparisons of England, Wales, and Scotland give datapoints that lie very close to the diagonal. The comparisons with N. Ireland show a few datapoints that differ significantly, with higher consumption of fresh potatos and lower consumption of fresh fruit. However, a holistic comparison is difficult.


```{r, echo=F}
pca <- prcomp( t(x) )
summary(pca)
```

Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.


```{r, echo=F}
plot(pca$x[ ,1], pca$x[ ,2], xlab="PC1", ylab="PC2", xlim=c(-300,500), pch=15)
text(pca$x[,1], pca$x[,2], colnames(x), adj=c(1,0))

```


Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r, echo=F}
plot(pca$x[ ,1], pca$x[ ,2], xlab="PC1", ylab="PC2", xlim=c(-300,500), pch=20, col="gray20")
text(pca$x[,1], pca$x[,2], colnames(x), adj=c(1,0), col=c("orange", "coral", "cyan", "green"))

```

```{r, echo=F}
plot(pca$x[ ,1], pca$x[ ,2], xlab="PC1", ylab="PC2", xlim=c(-300,500), pch=20, col="gray20")
text(pca$x[,1], pca$x[,2], colnames(x), adj=c(1,0), col=c("orange", "coral", "cyan", "green"))

```




