---
title: "Coronavirus Death Rates"
author: "Michael Overton"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This script analyzes infection data for the 2019 novel Coronavirus COVID-19 (2019-nCoV) epidemic. The raw data is pulled from the Johns Hopkins University Coronavirus repository.

The CSV file is available here https://github.com/RamiKrispin/coronavirus-csv
 
Read in data
```{r}

url <- "https://tinyurl.com/COVID-2019"
virus <- read.csv(url)
```

Q1) What is the total number of cases of COVID-19 worldwide?
```{r cars}

total_cases <- sum(virus$cases)
print(total_cases)
```
Q2) How many deaths are associated with COVID-19 worldwide, and what is the death rate?
```{r}
# Number of deaths
total_deaths <- sum(virus[virus$type == "death",]$cases)

# What is the overall death rate?
total_dr <- round(total_deaths/total_cases,4)

print(paste(total_dr*100,"%",sep=""))
```

Q4) What is the death rate in Mainland China?
```{r}
# What is the death rate in China?
china_cases <- sum(virus[virus$Country.Region == "Mainland China",]$cases)
china_deaths <- sum(virus[virus$Country.Region == "Mainland China" & virus$type == "death",]$cases)
china_dr <- round(china_deaths/china_cases,4)
print(paste(china_dr*100,"%",sep=""))
```

Q5. What are the death rates in Italy, Iran and the US?
```{r}
# What are the death rates in Italy, Iran, and the US?
italy_cases <- sum(virus[virus$Country.Region == "Italy",]$cases)
italy_deaths <- sum(virus[virus$Country.Region == "Italy" & virus$type == "death",]$cases)
italy_dr <- italy_deaths/italy_cases

iran_cases <- sum(virus[virus$Country.Region == "Iran",]$cases)
iran_deaths <- sum(virus[virus$Country.Region == "Iran" & virus$type == "death",]$cases)
iran_dr <- iran_deaths/iran_cases

us_cases <- sum(virus[virus$Country.Region == "US",]$cases)
us_deaths <- sum(virus[virus$Country.Region == "US" & virus$type == "death",]$cases)
us_dr <- us_deaths/us_cases

dr_df <- round(data.frame(Total=total_dr, US=us_dr, Iran=iran_dr, Italy=italy_dr), 4)

print("Death rate by country (%)")
print(dr_df*100)
```

Q6) Compare the death rates among the 10 countries with the highest number of cases.
```{r}
library(tidyverse)
wide_virus <- pivot_wider(virus, names_from = type,
  names_repair = "unique", values_from = cases, values_fill = list(cases = 0))

cases_by_country <- wide_virus %>%
  group_by(Country.Region) %>%
  summarise(total = sum(confirmed, death, recovered), deaths = sum(death)) %>%
  arrange(desc(total))

cases_by_country$rate <- cases_by_country$deaths/cases_by_country$total

library(ggplot2)

cases_by_country[1:10,] %>% 
  ggplot(aes(x=reorder(Country.Region, -rate), y=(rate*100))) + geom_col(fill="navyblue") + 
  xlab("\nCountry (total cases)") + ylab("Death Rate (%)\n") + geom_text(aes(label = total), 
            position = position_dodge(width = 0.9), vjust = -0.25, size=3) + theme_minimal()
```

